---
title: "Color Luminance Analysis"
author: "Dominika Tkaczyk"
date: "October 24, 2016"
output: github_document
---

```{r imports, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
library(gridExtra)

source("testSet.R")
source("evaluation.R")
source("models.R")
source("features.R")
```

# Introduction

In this document we analyze colors with respect to whether they are dark or light. From now on, we will call this
feature luminance (https://en.wikipedia.org/wiki/Luma_%28video%29). The analysis deals with the following main
problem: how can we automatically determine whether the color is dark or light?

The issue is not trivial for a couple of reasons:

  * a color is not a single number, for which we could use a single threshold, but rather a combination of a few
  components (for example red, green and blue light in case of RGB model), each of which may affect the luminance
  differently,
  * color's luminance is subjective and may differ depending on who is judging it,
  * even though there are colors for which the correct answer is obvious (eg. white or black), there are also a
  lot of colors in case of which the answer might be ambiguous (eg. pure green).
 
## Problem Statement

We will represent colors using RGB model. In this model a color $c$ is represented by a vector of three numbers 
$c = [c_R, c_G, c_B]$, where the red, green and blue intensities are normalized, that is 
$0 \leq c_R, c_G, c_B \leq 1$. In this model, white is represented by a vector $white = [1, 1, 1]$, black by 
$black = [0, 0, 0]$, blue will be $blue = [0, 0, 1]$, and so on.

Color's luminance will be represented by a single categorical variable with two levels: *dark* and *light*.

The problem analyzed in this document is the following: find a function $L: [0,1]^3 \rightarrow \{d,l\}$, which
for a given color determines its luminance.

# Data

A dedicated dataset was prepared manualy for the purpose of this analysis. First, 300 colors were generated by
randomly sampling from uniform distribution for each RGB components separately. Then the colors were labelled
manually as *dark* or *light*. For manual labelling we used scatter plots of green intensity against red intensity with points coloured by the "real" color of the data point and labelled with its number from the data set:

```{r buildingTestset, echo=FALSE}
set.seed(2690)
set <- createColorSet(size = 20)
visualizeColorSet(set)
```

# Exploratory Analysis

Before exploring our dataset, let's first divide it into train and test sets. All the analyses, plotting and
exploring will be done using only train part. In the end, the test set will be used to give a fair, reliable
estimate of the prediction's error.

```{r readingData}
data <- read.csv("data.csv")
set.seed(2690)
split <- 0.67
trainIndex <- createDataPartition(data$Lum, p = split, list = FALSE)
train <- data[trainIndex,]
row.names(train) <- NULL
test <- data[-trainIndex,]
row.names(test) <- NULL
```

Our train set contains `r nrow(train)` observation with `r ncol(train)` attributes describing each of them. Let's 
first see how the first few observation look like:

```{r headTrain}
head(train, n = 10)
```

**Red**, **Green** and **Blue** attributes are numeric, while **Lum** is a categorical variable with two levels: 
*L* and *D*:

```{r strTrain}
str(train)
```

Let's now plot the histograms of all the variables:

```{r histograms, echo=FALSE}
colCompHist <- function(col) {
    ggplot(data = train, aes(train[[col]])) +
        geom_histogram(aes(y =..count../sum(..count..)), binwidth = 0.05, col = "black", fill = "white") +
        labs(x = col, y = "Fraction")
}
rHist <- colCompHist("Red")
gHist <- colCompHist("Green")
bHist <- colCompHist("Blue")
lAggr <- aggregate(Red ~ Lum, data = train, length)
lHist <- ggplot(data = lAggr, aes(x = Lum, y = Red/nrow(train))) + 
    geom_bar(stat = "identity", col = "black", fill = "white") +
    labs(x = "Lum", y = "Fraction")
grid.arrange(rHist, gHist, bHist, lHist, ncol = 2)
```

The distributions of red, green and blue components seem fairly uniform (this is to be expected, based on how the
data set was generated). The dark and light classes are also well balanced, we have similar number of observations
in each class.

Let's also examine the boxplots of each variable divided by the luminance:

```{r boxplots, echo=FALSE}
rbp <- ggplot(train, aes(x = Lum, y = Red, fill = Lum)) + geom_boxplot()
gbp <- ggplot(train, aes(x = Lum, y = Green, fill = Lum)) + geom_boxplot()
bbp <- ggplot(train, aes(x = Lum, y = Blue, fill = Lum)) + geom_boxplot()
grid.arrange(rbp, gbp, bbp, ncol=2)
```

It can be clearly seen that color components affect the luminance differently:

  * green seems to keep the most information about the luminance; the means of green components for dark and light
  colors differ greatly, and the IQRs of the distributions do not overlap,
  * red seem to be less correlated with luminance; the means are still far apart, but the IQRs are closer to each
  other,
  * blue separated the dark and light colors the worst, the two distributions are pretty similar.

We can also perform statistical tests to confirm these observations more formally. We will use non-parametric
Kolmogorov–Smirnov test to compare two samples: red/green/blue values for dark and light colors. The p-value will
indicate how probable it is that the samples come from the same distibution.

```{r statTests}
ks.test(train[train$Lum=='D',]$Red, train[train$Lum=='L',]$Red)
ks.test(train[train$Lum=='D',]$Green, train[train$Lum=='L',]$Green)
ks.test(train[train$Lum=='D',]$Blue, train[train$Lum=='L',]$Blue)
```

As observed before, in the case of *Blue* we obtained a large p-value, suggesting that the samples are not that 
different. *Red* and *Green*, however, resulted in a very small p-values, so in these cases the distributions
indeed vary a lot.

Let's also examine the correlations between the variables:

```{r correlations}
cor(train[,1:3])
```

There seem to be no correlations between *Red*, *Green* and *Blue* componenets in our data set. Again, this is to
be expected considering how the data set was built.

Let's also plot variables against each other and color the points with luminance:

```{r pairs}
pairs(train, col=train$Lum)
```

The plot of *Red* vs. *Green* shows a fairly clear linear separation between the luminance classes. Similar trend
can be observed in the *Green* vs. *Blue* plot, but somewhat weaker: more points of different classes lie close
together in the middle area. The plot *Red* vs. *Blue* shows no pattern.

The initial analysis results in the following conclusions:

  * green components seem to be the most informative of the luminance, and blue - the least,
  * linear methods might be a good approach to the prediction problem, as the plots show the possibility of a 
  linear separation between dark and light classes.

# Evaluation

We will use the train portion of the dataset to evaluate and compare various models. Since luminance classes are
balanced, we will use accuracy as the main evaluation metric. A result for a single model is an array of 10 
accuracy values obtained by a 10-fold cross validation on the train set. In the first version we will compare the 
models using a mean of those values.

Function *evaluate* takes a model function and returns a vector of 10 accuracies obtained during a 10-fold cross
validation. A *model function* should take two parameters: a training and a testing set, and after learning the 
model using the training set it should output the predicted values for the testing set. In this setting, it will
be executed 10 times by the *evaluate* function with the arguments equal to 9/10 and the remaining 1/10 of the 
train set, respectively.

Example execution (this random model function ignores the training set):

```{r modelRandom}
modelRandom <- function(training, testing) {
    set.seed(2690)
    sample(c("D", "L"), nrow(testing), replace = TRUE)
}
cvSummary(evaluate(train, modelRandom))
```

# Prediction Models

## Baseline Models

Let's start with a few simple baseline models, which do not require any training at all. We will use the following
models:

  * **modelAlwaysLight** predicts *light* for all data points,
  * **modelHalfSum** predicts *light* when average color components intensity is greater than or equal 0.5, that
  is when $c_R + c_G + c_B \geq 1.5$
  * **modelStandardLuma** uses the standard luma definition for digital formats (see 
  https://en.wikipedia.org/wiki/Luma_%28video%29)

```{r baselineModels}
modelAlwaysLightResult <- evaluate(train, modelAlwaysLight)
modelHalfSumResult <- evaluate(train, modelHalfSum)
modelStandardLumaResult <- evaluate(train, modelStandardLuma)
```

Let's also compare out baseline models visually by plotting box plots of the distributions of the accuracies 
obtained by each model during cross validation:

```{r baselineVis}
results <- list(
    "Always light"  = modelAlwaysLightResult,
    "Half sum"      = modelHalfSumResult,
    "Standard luma" = modelStandardLumaResult
)
visualizeResults(results)
```

The following table summarizes baseline models' results. To compare the models more formally, we use statistical
tests to find out whether the differences between the results obtained in the cross validation are statistically
significant. Wilcoxon signed-rank test is used to compare the results and the models are ordered by the number of
statistical wins:

```{r baselineSummary, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE}
summarizeResults(results)
```

We can clearly see that standard luma equation gives very good results and outperforms other baseline methods.

## Learning Models

Now let's fit a few models to the dataset and see how they perform. We are using *caret* package which by default
performs a grid search in order to find the best parameters for a model. We will try the following learning 
methods:

  * Logistic Regression
  * Linear Discriminant Analysis and Quadratic Discriminant Analysis
  * Naive Bayes
  * Decision Tees and Random Forest
  * Support Vector Machines with linear, polynomial and radial kernel
  * K-Nearest Neighbors

We will also enhance our dataset with new features:

  * the squares of all RGB components: *SqRed*, *SqGreen*, *SqBlue*
  * the square roots of all RGB components: *SqRootRed*, *SqRootGreen*, *SqRootBlue*
  * multiplications of the components: *MultRedGreen*, *MultGreenBlue*, *MultRedBlue*, *MultRedGreenBlue*
  * ratios of the components: *RatioRedGreen*, *RatioGreenRed*, *RatioedBlue*, *RatioBlueRed*, *RatioBlueGreen*,
  *RatioGreenBlue*, *RatioRedGreenBlue*

Each model starts with a different base sets of the features and performs feature selection within each fold 
(Recursive Feature Elimination using caret-provided feature importance scores).

### Basic Features

These models use only the three basic features:

```{r modelsBasic, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache = TRUE}
train <- extractFeatures(train)

initResultDF <- function() {
    data.frame(basic = rep(0, 10), 
               mult  = rep(0, 10),
               ratio = rep(0, 10),
               sq    = rep(0, 10))
}
modelLogitResult     <- initResultDF()
modelLDAResult       <- initResultDF()
modelQDAResult       <- initResultDF()
modelNBResult        <- initResultDF()
modelTreeResult      <- initResultDF()
modelRFResult        <- initResultDF()
modelSVMLinearResult <- initResultDF()
modelSVMRadialResult <- initResultDF()
modelSVMPolyResult   <- initResultDF()
modelKNNResult       <- initResultDF()

log <- capture.output({

modelLogitResult$basic      <- evaluate(train, modelLogit,     features = basicFeatures)
modelLDAResult$basic        <- evaluate(train, modelLDA,       features = basicFeatures)
modelQDAResult$basic        <- evaluate(train, modelQDA,       features = basicFeatures)
modelNBResult$basic         <- evaluate(train, modelNB,        features = basicFeatures)
modelTreeResult$basic       <- evaluate(train, modelTree,      features = basicFeatures)
modelRFResult$basic         <- evaluate(train, modelRF,        features = basicFeatures)
modelSVMLinearResult$basic  <- evaluate(train, modelSVMLinear, features = basicFeatures)
modelSVMRadialResult$basic  <- evaluate(train, modelSVMRadial, features = basicFeatures)
modelSVMPolyResult$basic    <- evaluate(train, modelSVMPoly,   features = basicFeatures)
modelKNNResult$basic        <- evaluate(train, modelKNN,       features = basicFeatures)

})

results <- list(
    "Logistic regression" = modelLogitResult$basic,
    "LDA"                 = modelLDAResult$basic,
    "QDA"                 = modelQDAResult$basic,
    "Naive Bayes"         = modelNBResult$basic,
    "Decision tree"       = modelTreeResult$basic,
    "Random forest"       = modelRFResult$basic,
    "KNN"                 = modelKNNResult$basic,
    "SVM linear"          = modelSVMLinearResult$basic,
    "SVM radial"          = modelSVMRadialResult$basic,
    "SVM poly"            = modelSVMPolyResult$basic
)
visualizeResults(results)
summarizeResults(results)
```

### Basic Features + Multiplication Features

These models use three basic features and features based on multiplication:

```{r modelsMult, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=TRUE}
fs <- c(basicFeatures, multFeatures)

log <- capture.output({
    
modelLogitResult$mult     <- evaluate(train, modelLogit,     features = fs)
modelLDAResult$mult       <- evaluate(train, modelLDA,       features = fs)
modelQDAResult$mult       <- evaluate(train, modelQDA,       features = fs)
modelNBResult$mult        <- evaluate(train, modelNB,        features = fs)
modelTreeResult$mult      <- evaluate(train, modelTree,      features = fs)
modelRFResult$mult        <- evaluate(train, modelRF,        features = fs)
modelSVMLinearResult$mult <- evaluate(train, modelSVMLinear, features = fs)
modelSVMRadialResult$mult <- evaluate(train, modelSVMRadial, features = fs)
modelSVMPolyResult$mult   <- evaluate(train, modelSVMPoly,   features = fs)
modelKNNResult$mult       <- evaluate(train, modelKNN,       features = fs)

})

results <- list(
    "Logistic regression" = modelLogitResult$mult,
    "LDA"                 = modelLDAResult$mult,
    "QDA"                 = modelQDAResult$mult,
    "Naive Bayes"         = modelNBResult$mult,
    "Decision tree"       = modelTreeResult$mult,
    "Random forest"       = modelRFResult$mult,
    "KNN"                 = modelKNNResult$mult,
    "SVM linear"          = modelSVMLinearResult$mult,
    "SVM radial"          = modelSVMRadialResult$mult,
    "SVM poly"            = modelSVMPolyResult$mult
)
visualizeResults(results)
summarizeResults(results)
```

### Basic Features + Multiplication Features + Ratio Features

These models use three basic features, features based on multiplication and ratio-based features:

```{r modelsRatio, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=TRUE}
fs <- c(basicFeatures, multFeatures, ratioFeatures)

log <- capture.output({

modelLogitResult$ratio     <- evaluate(train, modelLogit,     features = fs)
modelLDAResult$ratio       <- evaluate(train, modelLDA,       features = fs)
modelQDAResult$ratio       <- evaluate(train, modelQDA,       features = fs)
modelNBResult$ratio        <- evaluate(train, modelNB,        features = fs)
modelTreeResult$ratio      <- evaluate(train, modelTree,      features = fs)
modelRFResult$ratio        <- evaluate(train, modelRF,        features = fs)
modelSVMLinearResult$ratio <- evaluate(train, modelSVMLinear, features = fs)
modelSVMRadialResult$ratio <- evaluate(train, modelSVMRadial, features = fs)
modelSVMPolyResult$ratio   <- evaluate(train, modelSVMPoly,   features = fs)
modelKNNResult$ratio       <- evaluate(train, modelKNN,       features = fs)

})

results <- list(
    "Logistic regression" = modelLogitResult$ratio,
    "LDA"                 = modelLDAResult$ratio,
    "QDA"                 = modelQDAResult$ratio,
    "Naive Bayes"         = modelNBResult$ratio,
    "Decision tree"       = modelTreeResult$ratio,
    "Random forest"       = modelRFResult$ratio,
    "KNN"                 = modelKNNResult$ratio,
    "SVM linear"          = modelSVMLinearResult$ratio,
    "SVM radial"          = modelSVMRadialResult$ratio,
    "SVM poly"            = modelSVMPolyResult$ratio
)
visualizeResults(results)
summarizeResults(results)
```

### Basic Features + Multiplication Features + Ratio Features + Square Features

These models use three basic features and all additional features:

```{r modelsSq, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=TRUE}
fs <- c(basicFeatures, multFeatures, ratioFeatures, sqFeatures)

log <- capture.output({

modelLogitResult$sq     <- evaluate(train, modelLogit,     features = fs)
modelLDAResult$sq       <- evaluate(train, modelLDA,       features = fs)
modelQDAResult$sq       <- evaluate(train, modelQDA,       features = fs)
modelNBResult$sq        <- evaluate(train, modelNB,        features = fs)
modelTreeResult$sq      <- evaluate(train, modelTree,      features = fs)
modelRFResult$sq        <- evaluate(train, modelRF,        features = fs)
modelSVMLinearResult$sq <- evaluate(train, modelSVMLinear, features = fs)
modelSVMRadialResult$sq <- evaluate(train, modelSVMRadial, features = fs)
modelSVMPolyResult$sq   <- evaluate(train, modelSVMPoly,   features = fs)
modelKNNResult$sq       <- evaluate(train, modelKNN,       features = fs)

})

results <- list(
    "Logistic regression" = modelLogitResult$sq,
    "LDA"                 = modelLDAResult$sq,
    "QDA"                 = modelQDAResult$sq,
    "Naive Bayes"         = modelNBResult$sq,
    "Decision tree"       = modelTreeResult$sq,
    "Random forest"       = modelRFResult$sq,
    "KNN"                 = modelKNNResult$sq,
    "SVM linear"          = modelSVMLinearResult$sq,
    "SVM radial"          = modelSVMRadialResult$sq,
    "SVM poly"            = modelSVMPolyResult$sq
)
visualizeResults(results)
summarizeResults(results)
```

### Model Comparison

Now we will compare all the models built so far, including the baseline models:

```{r resultsAll, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=TRUE}
results <- list(
    "Always light"                = modelAlwaysLightResult,
    "Half sum"                    = modelHalfSumResult,
    "Standard luma"               = modelStandardLumaResult,
    
    "Logistic regression (basic)" = modelLogitResult$basic,
    "LDA (basic)"                 = modelLDAResult$basic,
    "QDA (basic)"                 = modelQDAResult$basic,
    "Naive Bayes (basic)"         = modelNBResult$basic,
    "Decision tree (basic)"       = modelTreeResult$basic,
    "Random forest (basic)"       = modelRFResult$basic,
    "KNN (basic)"                 = modelKNNResult$basic,
    "SVM linear (basic)"          = modelSVMLinearResult$basic,
    "SVM radial (basic)"          = modelSVMRadialResult$basic,
    "SVM poly (basic)"            = modelSVMPolyResult$basic,
    
    "Logistic regression (mult)"  = modelLogitResult$mult,
    "LDA (mult)"                  = modelLDAResult$mult,
    "QDA (mult)"                  = modelQDAResult$mult,
    "Naive Bayes (mult)"          = modelNBResult$mult,
    "Decision tree (mult)"        = modelTreeResult$mult,
    "Random forest (mult)"        = modelRFResult$mult,
    "KNN (mult)"                  = modelKNNResult$mult,
    "SVM linear (mult)"           = modelSVMLinearResult$mult,
    "SVM radial (mult)"           = modelSVMRadialResult$mult,
    "SVM poly (mult)"             = modelSVMPolyResult$mult,
    
    "Logistic regression (ratio)" = modelLogitResult$ratio,
    "LDA (ratio)"                 = modelLDAResult$ratio,
    "QDA (ratio)"                 = modelQDAResult$ratio,
    "Naive Bayes (ratio)"         = modelNBResult$ratio,
    "Decision tree (ratio)"       = modelTreeResult$ratio,
    "Random forest (ratio)"       = modelRFResult$ratio,
    "KNN (ratio)"                 = modelKNNResult$ratio,
    "SVM linear (ratio)"          = modelSVMLinearResult$ratio,
    "SVM radial (ratio)"          = modelSVMRadialResult$ratio,
    "SVM poly (ratio)"            = modelSVMPolyResult$ratio,
    
    "Logistic regression (sq)"    = modelLogitResult$sq,
    "LDA (sq)"                    = modelLDAResult$sq,
    "QDA (sq)"                    = modelQDAResult$sq,
    "Naive Bayes (sq)"            = modelNBResult$sq,
    "Decision tree (sq)"          = modelTreeResult$sq,
    "Random forest (sq)"          = modelRFResult$sq,
    "KNN (sq)"                    = modelKNNResult$sq,
    "SVM linear (sq)"             = modelSVMLinearResult$sq,
    "SVM radial (sq)"             = modelSVMRadialResult$sq,
    "SVM poly (sq)"               = modelSVMPolyResult$sq
)
summary <- summarizeResults(results)
summary[,2:5] <- round(summary[,2:5], 5)
summary
```

This is a visualizaton of 10 best models:

```{r visAll, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=TRUE}
summaryTop10 <- summary[1:10,]$Method
visualizeResults(results[names(results) %in% summaryTop10])
```

It seems that SVM model with radial kernel and basic combined with multiplication-based features
achieves the best results, at the same time beating the most other models in statistical tests.
Let's make one final evaluation of the winning model using the held-out dataset:

```{r finalEval, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, cache=TRUE}
test <- extractFeatures(test)
testResults <- trainAndPredict(train, test, "svmRadial", 
                               features = c(basicFeatures, multFeatures))
cm <- confusionMatrix(test$Lum, testResults)
cm
meanStandardLuma <- mean(evaluate(train, modelStandardLuma))
meanSVMRadial <- mean(evaluate(train, modelSVMRadial, features = c(basicFeatures, multFeatures)))
```

# Conclusions

In this document we examined the possibility of automated classifying a given color as *light* or
*dark* based on their RGB components. The most important findings of the analysis include:

  * green component is the most correlated with the target luminance, and blue component - the least
  * a simple linear model based on the standard luma definition for digital formats performs well,
  acheiving the mean accuracy of `r meanStandardLuma` on our training set
  * the best model found is SVM model with radial kernel and basic combined with
  multiplication-based features, achieving the mean accuracy of `r meanSVMRadial` on our training
  set
  * the accuracy of the best model on our held-out set (and thus our best estimate for the accuracy
  of the chosen solution is `r cm$overall[["Accuracy"]]`
